
\subsection{Background}

The study of income distributions has been a cornerstone of economic research. A long-standing empirical regularity in the literature is that income distributions tend to approximate a log-normal density \citep{aitchison1957lognormal}. This stylized fact is supported by both theoretical and empirical foundations. At its core, log-normality arises from the multiplicative interaction of economic factors -- such as human capital, local labor market conditions, and productivity -- which determine individual incomes \citep{neal2000theories}. Gibrat's Law of Proportionate Effect \citep{gibrat1931inegalites} formalizes the theoretical basis for log-normality through multiplicative random growth, showing that when the growth rate of a variable is independent of its initial size and the logarithm of the growth rate is independent and identically distributed over time with finite variance, the resulting distribution tends to converge to log-normality. 

More concretely, the multiplicative processes through which log-normality arises operate most strongly within demographically homogeneous groups where individuals are subject to similar economic shocks \citep{weiss1972, aitchison1957lognormal, battistin2009consumption}. This methodology note relates to a larger body of research providing empirical support for this pattern, showing that a mixture of log-normals is particularly suitable for modeling income in fairly homogeneous sub-populations while at the same time accounting for between-group heterogeneity \citep{flachaire2007, lubrano2016}. This finding is particularly relevant at higher levels of aggregation, where finite mixtures of log-normals provide substantial improvements in fit compared to single log-normal specifications \citep{gardini2022}.

The literature of neighborhood effects offers a complementary explanation for why such homogeneity in income-generating processes exists at local levels. In particular, neighborhood effects reinforce homogeneity within local populations through contextual influences and endogenous spillovers, which emerge from behavioral interactions within the neighborhood, such as peer influences or social norms \citep{manski1993identification, durlauf1996neighborhoods}. These mechanisms create reinforcing feedback loops that generate correlated income trajectories within neighborhoods -- for instance, through shared information about job opportunities, similar human capital accumulation patterns, or common responses to local economic shocks.  The resulting interaction structures generate strong within-neighborhood homogeneity while maintaining between-neighborhood heterogeneity \citep{durlauf2004neighborhood}. This pattern provides a rationale for estimating aggregate income distributions as mixtures of local-level distributions.

However, while providing a good approximation for much of the income distribution, the log-normal approximation exhibits systematic deviations in the tails. Since Pareto's \citeyearpar{pareto1896cours} seminal work, research has shown that top incomes follow a power law rather than a log-normal decay \citep{gabaix2016power}. Similarly, studies using detailed administrative tax data have documented that the upper tail of income distributions across countries and time periods are better described by a Pareto distribution \citep{atkinson2011top}. This departure from log-normality at high incomes is important for accurately measuring top income inequality, but modeling the precise behavior of the upper tail is beyond the scope of this note.

% (e.g., average neighborhood income or quality of local services)

%Relatedly, a more recent literature on place effects has documented that the environment where children grow up has substantial causal impacts on their economic outcomes later in life \cite{chynplace}. Using experimental variation from the Moving to Opportunity program, \cite{chetty2016effects} show that childhood exposure to better neighborhoods significantly increases college attendance and earnings in adulthood.

\subsection{Empirical approach}

I estimate the national income distribution by exploiting the granular structure of census tract-level data and the theoretical properties of income distributions. My approach leverages the fact that income distributions within small geographic units tend to follow log-normal distributions more closely than in larger areas, and approximates the aggregate distribution using a mixture of these local-level distributions.

Consider a census tract $j$ with an observed \emph{median} income $\tilde m_j$ and Gini coefficient $G_j$. Under the log-normality assumption, if income $Y$ follows a log-normal distribution with parameters $(\nu_j, \sigma_j^2)$, then

\begin{equation}
\ln(Y) \sim N(\nu_j, \sigma_j^2).
\end{equation}

The mapping between these parameters and observed statistics is:

\begin{equation}
\tilde m_j = \exp(\nu_j),
\end{equation}

\begin{equation}
G_j = 2\Phi\left(\frac{\sigma_j}{\sqrt{2}}\right) - 1,
\end{equation}

where $\Phi(\cdot)$ is the standard normal cumulative distribution function. From the observed Gini coefficient, I recover $\sigma_j$ as

\begin{equation}
\sigma_j = \sqrt{2}\,\Phi^{-1}\!\left(\frac{G_j + 1}{2}\right),
\end{equation}

and from the median I identify $\nu_j=\log \tilde m_j$. When the tract median is unavailable but the (equivalised) \emph{mean} income $\mu_j$ is observed, I use the exact log-normal relationship

\begin{equation}
\mu_j = \exp\!\left(\nu_j + \frac{\sigma_j^2}{2}\right)
\quad\Rightarrow\quad
\nu_j = \log \mu_j - \frac{\sigma_j^2}{2}.
\end{equation}

I then estimate the national distribution as a population-weighted mixture of these local log-normal distributions. For a given income level $y$, the density is

\begin{equation}
f(y) = \sum_{j=1}^{J} w_j\, f_j(y \mid \nu_j,\sigma_j^2),
\end{equation}

where $w_j$ is tract $j$'s population share and $f_j(\cdot \mid \nu_j,\sigma_j^2)$ is the log-normal density. Percentiles of the national distribution are obtained by solving

\begin{equation}
p = F(q_p) = \sum_{j=1}^{J} w_j \,\Phi\!\left(\frac{\ln(q_p) - \nu_j}{\sigma_j}\right),
\end{equation}

where $q_p$ is the $p^{th}$ percentile and $F(\cdot)$ is the cumulative distribution function of the mixture. I use numerical root-finding methods to find the value of $q_p$ that satisfies $F(q_p) - p = 0$ within a bounded interval.

%This approach has several advantages. First, it respects the theoretical properties of income distributions at local levels where populations are relatively homogeneous. Second, it preserves tract-level moments while allowing for heterogeneity across tracts. Third, it provides a flexible framework that can be adapted and extended to calculate any distributional statistic of interest.

\paragraph{Missing outcomes} For census tracts with missing Gini coefficients (approximately 5.5\% of the sample), I estimate them using a machine learning approach. Specifically, I train an XGBoost model using the following demographic predictors: dependency ratio (ratio of population under 18 and over 65 to working-age population), mean age, percentage of single-person households, logged mean equivalised income, mean household size, population size, and province fixed effects. Model performance is validated using a 5-fold cross-validation procedure, and is trained using default hyperparameters without grid search or additional tuning. For comparison, I also estimate a baseline OLS model using the same set of predictors. Out-of-sample prediction errors under both approaches are shown in Table \ref{perf}. Based on the cross-validation results, the XGBoost model demonstrates superior predictive performance compared to a baseline OLS model, as illustrated by a 6\% lower RMSE.

\begin{table}[H]
\centering
\resizebox{0.7\textwidth}{!}{% 
\begin{threeparttable}[H]
\onehalfspacing
\centering
\captionsetup{justification=centering} 
\caption{Performance comparison between OLS and XGBoost models}\label{perf}
\begin{tabular}{lccc}
\hline
Model    & RMSE & MAE & MAPE (\%) \\
\hline
OLS                & 2.98          & 2.29         & 8.02              \\
XGBoost (CV)       & 2.80          & 2.14         & 7.51              \\
\hline
Relative improvement over OLS (\%) & 6.04\% & 6.55\% & 6.36\% \\
\hline
\end{tabular}
\begin{tablenotes}[para,flushleft]
\scriptsize 
\textbf{Notes}: This table compares the out-of-sample performance of two predictive models -- OLS and XGBoost -- on predicting missing Gini coefficients using demographic predictors. RMSE represents the root mean square error, MAE is the mean absolute error, and MAPE is the mean absolute percentage error. Relative improvements in each metric are calculated as the percentage reduction of the error metric when using XGBoost compared to OLS. Results are based on a 5-fold cross-validation procedure for the XGBoost model.
\end{tablenotes}
\end{threeparttable}
}
\end{table}

To validate the log-normality assumption at the tract level, I compare observed distributional statistics with their theoretical counterparts under log-normality. For each tract $j$, I recover $(\nu_j,\sigma_j)$ using the observed median and Gini coefficient. I then compute the following predicted statistics:

\begin{equation}
\widehat{\text{Mean}}_j \;=\; \exp\!\left(\nu_j + \frac{\sigma_j^2}{2}\right),
\end{equation}

\begin{equation}
\widehat{\text{P80/P20}_j} \;=\; \exp\!\left(\sigma_j \,[\Phi^{-1}(0.8) - \Phi^{-1}(0.2)]\right).
\end{equation}

To assess the fit, I estimate OLS regressions of the form

\begin{equation}
\hat{y}_j = \alpha + \beta\, y_j + \epsilon_j,
\end{equation}

where $\hat{y}_j$ is the predicted value under log-normality and $y_j$ is the observed value (either the tract mean income or the P80/P20 ratio).
